Marlowe: financial contracts on blockchain

### 2018-05-17 tex/paper.tex
some compacting
added textual description to Sec 2.2, and reformatted Sec 2.3

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.6\textwidth]{pix/context.png}
\caption{The context for a contract}
\label{fig:context}
\end{center}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.8\textwidth]{pix/step-type.png}
\caption{The \haskellinline{step} function}
\label{fig:step-function}
\end{center}
\end{figure}

@@

\paragraph{Computations}

@@

Execution of a contract will involve multiple blocks, with multiple steps in each block. The computation at a single block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls \haskellinline{step} repeatedly
until it is quiescent. In addition to 
calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step. 
The functions \haskellinline{stepAll} and \haskellinline{stepBlock} have the same type as \haskellinline{step} itself.

@@

\begin{haskellcode}
data Contract =
   Null |
   CommitCash IdentCC Person Money Timeout Timeout Contract Contract |  
   RedeemCC IdentCC Contract |
   Pay IdentPay Person Person Money Timeout Contract |  
   Both Contract Contract |
   Choice Observation Contract Contract |
   When Observation Timeout Contract Contract   
   \end{haskellcode}
%Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be 
%archived in the blockchain? I don't see how your design will support doing so.}\footnote{Yes, we assume that it is 
%part of the infrastructure of this language to record the values of *observables* that are used. However, we have 
%chosen no to make it part of the language itself.}

\subsection{The \haskellinline{step} function}

The constructors are interpreted like this:

\begin{itemize}

\item \haskellinline{Null} is the null contract; it will always be quiescent:

\smallskip
\begin{haskellcode}
step _ st Null _ = (st, Null, [])
\end{haskellcode}
\smallskip

\item \haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value 
\haskellinline{v} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC}, 
\item
or the timeout \haskellinline{t1} is exceed:
\end{itemize}

\smallskip
\begin{haskellcode}
step 
  commits 
  st 
  c@(CommitCash ident person val start_timeout end_timeout con1 con2) 
  os
  | cexe || cexs = (st {sc = ust}, con2, [])
  | Set.member (CC ident person cval end_timeout) (cc commits)
        = (st {sc = ust}, con1, [SuccessfulCommit ident person cval])
  | otherwise = (st, c, [])
  where ccs = sc st
        cexs = expired (blockNumber os) start_timeout
        cexe = expired (blockNumber os) end_timeout
        cns = (person, if cexe || cexs 
                          then ManuallyRedeemed 
                          else NotRedeemed cval end_timeout)
        ust = Map.insert ident cns ccs
        cval = evalMoney st val
\end{haskellcode}
\smallskip
In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
\haskellinline{k1}; in the second case no action is generated and the contract continues as 
\haskellinline{k2}. While neither case holds, the contract is quiescent, waiting for the cash to be committed.

@@

\item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent 
funds in that commitment; the contract then continues as \haskellinline{k}, and the action 
\haskellinline{CommitRedeemed} is produced.

\smallskip
\begin{haskellcode}
step commits st c@(RedeemCC ident con) _ =
    case Map.lookup ident ccs of
      Just (person, NotRedeemed val _) ->
        let newstate = 
                st {sc = Map.insert ident (person, ManuallyRedeemed) ccs} in
        if Set.member (RC ident person val) (rc commits)
        then (newstate, con, [CommitRedeemed ident person val])
        else (st, c, [])
      Just (person, ManuallyRedeemed) ->
        (st, con, [DuplicateRedeem ident person])
      Nothing -> (st,c,[])
    where
        ccs = sc st
\end{haskellcode}
\smallskip

Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  
\haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.

@@

\item \haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
\haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
\haskellinline{t}. The contract continues as \haskellinline{k}.

\smallskip
\begin{haskellcode}
step inp st c@(Pay idpay from to val expi con) os
  | expired (blockNumber os) expi = (st, con, [ExpiredPay idpay from to cval])
  | right_claim =
    if committed st from bn >= cval
      then (newstate, con, [SuccessfulPay idpay from to cval])
      else (st, con, [FailedPay idpay from to cval])
  | otherwise = (st, c, [])
  where
    cval = evalMoney st val
    newstate = stateUpdate st from to bn cval
    bn = blockNumber os
    right_claim =
      case Map.lookup (idpay, to) (rp inp) of
        Just claimed_val -> claimed_val == cval
        Nothing -> False
\end{haskellcode}
\smallskip

By `available' we mean that sufficient commitments have 
been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash 
commitments made by \haskellinline{p1} that expire the earliest.

@@

\item \haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}.

\smallskip
\begin{haskellcode}
step comms st (Both con1 con2) os =
    (st2, result, ac1 ++ ac2)
    where
        result | res1 == Null = res2
               | res2 == Null = res1
               | otherwise = Both res1 res2
        (st1,res1,ac1) = step comms st con1 os
        (st2,res2,ac2) = step comms st1 con2 os
\end{haskellcode}
\smallskip

Because the model is stateful and produces output actions, to make a step, it is necessary to execute a single step of 
each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then 
\haskellinline{k2}.

\item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
(Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is 
\haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.

\smallskip
\begin{haskellcode}
step _ st (Choice obs conT conF) os =
    if interpretObs st obs os
        then (st,conT,[])
        else (st,conF,[])
\end{haskellcode}
\smallskip

\item \haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
\haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout. In case 
the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and 
\haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}. Otherwise the contract is 
quiescent.

\smallskip
\begin{haskellcode}
step _ st (When obs expi con con2) os
  | expired (blockNumber os) expi = (st,con2,[])
  | interpretObs st obs os = (st,con,[])
  | otherwise = (st, When obs expi con con2, [])
  \end{haskellcode}
\smallskip

\end{itemize}
We look next at an example of Marlowe in action.

@@

Reflecting on the example, we can see that using Haskell definitions has made the contract substantially more 
comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression 
\haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause, 
\begin{haskellcode}
escrow = ...
                    (When (OrObs (two_chose 1 2 3 0)
                                 chose_refund)
                          90
                          (Choice chose_refund
                           ...
         ...
         where 
         chose_refund = two_chose 1 2 3 1                 
\end{haskellcode}
then the repeated computation of the expression will be avoided.

In the next section, we cover the online contract visualisation and interaction tool for Marlowe.

@@

\begin{figure}[t]
\begin{centering}
\includegraphics[width=1.1\textwidth]{pix/screenshot1}
\par\end{centering}
\caption{\label{fig:full-screenshot-demo}Screenshot of Meadow tool
simulating the incentivised deposit contract.}
\end{figure}

@@

\begin{figure}[t]
\centering{}\includegraphics[scale=0.52]{pix/detail}\caption{\label{fig:detail-of-interface}Detail of the interface for 
contract
execution simulation.}
\end{figure}

>>>>

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.6\textwidth]{pix/context.png}
\caption{The context for a contract}
\label{fig:context}
\end{center}
\vspace*{-8mm}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.8\textwidth]{pix/step-type.png}
\caption{The \haskellinline{step} function}
\label{fig:step-function}
\end{center}
\vspace*{-8mm}
\end{figure}

@@

\paragraph{Computation}

@@

Execution of a contract will involve multiple blocks, with multiple steps in each block. The computation at a single block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls \haskellinline{step} repeatedly
until it is quiescent.

In addition to 
calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step. 
The functions \haskellinline{stepAll} and \haskellinline{stepBlock} have the same type as \haskellinline{step} itself.

@@

\begin{haskellcode}
data Contract =
   Null |
   CommitCash IdentCC Person Money Timeout Timeout Contract Contract |  
   RedeemCC IdentCC Contract |
   Pay IdentPay Person Person Money Timeout Contract |  
   Both Contract Contract |
   Choice Observation Contract Contract |
   When Observation Timeout Contract Contract   
   \end{haskellcode}
Informally, this type provides a \haskellinline{Null} contract, which does nothing. The next three constructs form contracts that do something, and then continue according to another contract (which is one of the components of the original contract). \haskellinline{CommitCash} will wait for a participant to make a commitment,  
\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment to be made to another participant.

The remaining constructors form composite contracts from simpler components: \haskellinline{Both} has the behaviour of both its components, \haskellinline{Choice} chooses between two contracts on the basis of an observation, and \haskellinline{When} is quiescent until a condition -- i.e.\ an \haskellinline{Observation} -- becomes true.

What many of the contracts also have are timeouts that also determine their behaviour. 
\subsection{The \haskellinline{step} function}

In this section we explain the detailed behaviour of \haskellinline{Contracts} through a discussion of the \haskellinline{step} function.

\smallskip
\noindent
\haskellinline{Null} is the null contract; it will always be quiescent:

\begin{haskellcode}
step _ st Null _ = (st, Null, [])
\end{haskellcode}

\smallskip
\noindent
\haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value 
\haskellinline{v} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC}, 
\item
or the timeout \haskellinline{t1} is exceed:
\end{itemize}

\begin{haskellcode}
step 
  commits 
  st 
  c@(CommitCash ident person val start_timeout end_timeout con1 con2) 
  os
  | cexe || cexs = (st {sc = ust}, con2, [])
  | Set.member (CC ident person cval end_timeout) (cc commits)
        = (st {sc = ust}, con1, [SuccessfulCommit ident person cval])
  | otherwise = (st, c, [])
  where ccs = sc st
        cexs = expired (blockNumber os) start_timeout
        cexe = expired (blockNumber os) end_timeout
        cns = (person, if cexe || cexs 
                          then ManuallyRedeemed 
                          else NotRedeemed cval end_timeout)
        ust = Map.insert ident cns ccs
        cval = evalMoney st val
\end{haskellcode}
In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
\haskellinline{k1}; in the second case no action is generated and the contract continues as 
\haskellinline{k2}. While neither case holds, the contract is quiescent, waiting for the cash to be committed.

@@

\medskip
\noindent
\haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent 
funds in that commitment; the contract then continues as \haskellinline{k}, and the action 
\haskellinline{CommitRedeemed} is produced.

\begin{haskellcode}
step commits st c@(RedeemCC ident con) _ =
    case Map.lookup ident ccs of
      Just (person, NotRedeemed val _) ->
        let newstate = 
                st {sc = Map.insert ident (person, ManuallyRedeemed) ccs} in
        if Set.member (RC ident person val) (rc commits)
        then (newstate, con, [CommitRedeemed ident person val])
        else (st, c, [])
      Just (person, ManuallyRedeemed) ->
        (st, con, [DuplicateRedeem ident person])
      Nothing -> (st,c,[])
    where
        ccs = sc st
\end{haskellcode}
Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  
\haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.

@@

\medskip
\noindent
\haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
\haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
\haskellinline{t}. The contract continues as \haskellinline{k}.

\begin{haskellcode}
step inp st c@(Pay idpay from to val expi con) os
  | expired (blockNumber os) expi = (st, con, [ExpiredPay idpay from to cval])
  | right_claim =
    if committed st from bn >= cval
      then (newstate, con, [SuccessfulPay idpay from to cval])
      else (st, con, [FailedPay idpay from to cval])
  | otherwise = (st, c, [])
  where
    cval = evalMoney st val
    newstate = stateUpdate st from to bn cval
    bn = blockNumber os
    right_claim =
      case Map.lookup (idpay, to) (rp inp) of
        Just claimed_val -> claimed_val == cval
        Nothing -> False
\end{haskellcode}
By `available' we mean that sufficient commitments have 
been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash 
commitments made by \haskellinline{p1} that expire the earliest.

@@

\medskip
\noindent
\haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}. 
\begin{haskellcode}
step comms st (Both con1 con2) os =
    (st2, result, ac1 ++ ac2)
    where
        result | res1 == Null = res2
               | res2 == Null = res1
               | otherwise = Both res1 res2
        (st1,res1,ac1) = step comms st con1 os
        (st2,res2,ac2) = step comms st1 con2 os
\end{haskellcode}
Because the model is stateful and produces output actions, to make a step, it is necessary to execute a single step of 
each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then 
\haskellinline{k2}.

\medskip
\noindent
\haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
(Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is 
\haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.

\begin{haskellcode}
step _ st (Choice obs conT conF) os =
    if interpretObs st obs os
        then (st,conT,[])
        else (st,conF,[])
\end{haskellcode}

\medskip
\noindent
\haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
\haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout. In case 
the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and 
\haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}. Otherwise the contract is 
quiescent.

\begin{haskellcode}
step _ st (When obs expi con con2) os
  | expired (blockNumber os) expi = (st,con2,[])
  | interpretObs st obs os = (st,con,[])
  | otherwise = (st, When obs expi con con2, [])
  \end{haskellcode}
We look next at an example of Marlowe in action.

@@

Reflecting on the example, we can see that using Haskell definitions has made the contract substantially more 
comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression 
\haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause, 
\begin{haskellcode}
escrow = ...        (When (OrObs (two_chose 1 2 3 0)
                                 chose_refund)
                          90
                          (Choice chose_refund ...
                           ...
         where chose_refund = two_chose 1 2 3 1                 
\end{haskellcode}
then the repeated computation of the expression will be avoided.
In the next section, we cover the online contract visualisation and interaction tool for Marlowe.

@@

\begin{figure}[t]
\begin{centering}
\includegraphics[width=1.1\textwidth]{pix/screenshot1}
\par\end{centering}
\caption{\label{fig:full-screenshot-demo}Screenshot of Meadow tool
simulating the incentivised deposit contract.}
\vspace*{-8mm}
\end{figure}

@@

\begin{figure}[t]
\centering{}\includegraphics[scale=0.52]{pix/detail}\caption{\label{fig:detail-of-interface}Detail of the interface for 
contract
execution simulation.}
\vspace*{-5mm}
\end{figure}
